---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rTPI

<!-- badges: start -->
[![R-CMD-check](https://github.com/MacroEcoMatt/rTPI/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/MacroEcoMatt/rTPI/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of rTPI is to provide a set of tools for calculating 
Thermal Position and Aridity Position Indices (TPI and API). These indices are 
useful for evaluating the relative impact of temperature and aridity values 
given species specific thermal tolerance or aridity tolerance limits. 

You provide environmental, or location data with genus species binomial names, 
and are returned a dataframe with TPI or API values. These values are scaled
temperature and aridity values that are specific to each species niche
tolerance. Typically, values will fall between 0 (lower niche limits) and 1
(upper niche limit). However, values can exceed species tolerance limits and 
fall outside of 0 and 1, representing extreme environmental conditions given
species specific tolerances.

## Installation

You can install the development version of rTPI by downloading the package
directly and installing from the .zip file or like so:

``` r
# devtools::install_github("MacroEcoMatt/rTPI")
```

##Functions include:
binomial_check - cross check submitted species binomial (eg. Melospiza melodia)
against species binomials in the TPI/API dataset

syn_check - if submitted binomial does not match package binomial records, this
function will check user submitted binomial against known species synonyms and 
return matching TPI/API binomials

species_limits - used to extract a dataframe containint the TPI/API information
for submitted list of species

highertaxa_limits - used to extract a dataframe containint the TPI/API 
information for submitted list of higher taxa groups 
(e.g taxa_code = "clas", taxa_list = "Aves)

tpi - calculates TPI and prodcues a dataframe of values based on user submitted
data. Requires species binomial, and temperature data in degrees celcius

api - calculates API and prodcues a dataframe of values based on user submitted
data. Requires species binomial, and Aridity Index values. SEE get_env_vars.

```{r example}
library(rTPI)
sp_list <- c("Poecile atricapillus","Melospiza melodia")

##binomial_check
sp_check_df <- binomial_check(sp_list)

##syn_check
Compared_df <- syn_check(sp_list)

##species_limits
sp_limits_data <- species_limits(sp_list, niche_limit = NULL, month_list = NULL,
                           yr_avg = FALSE)

##highertaxa_limits
Aves_TPI <- highertaxa_limits(taxa_code = "class", taxa_list="Aves", 
                              niche_limit = "tpi", month_list="Jan", 
                              yr_avg=TRUE)

##TPI AND API
species_data <- data.frame(
                Binomial = c("Poecile atricapillus", "Poecile atricapillus"),
                Month = c(1,2),
                TMax = c(5,10),
                TMin = c(-2,-1),
                Tm = c(0,3),
                AMin = c(0,0.5),
                AMax = c(2,3),
                Ar = c(1,1.1)
                )

tpi_Poec_atr <- tpi(species_data, tmp_var = NULL, use_year = FALSE, flag_sp = FALSE,
                flag_month = FALSE, flag_tmp = FALSE)

api_Poec_atr <- api(species_data, ar_var = NULL, use_year = FALSE, flag_sp = FALSE,
                flag_month = FALSE, flag_ar = FALSE)
```

#Formatting:
This is important for the get_env_vars, tpi, and api functions as users need to
submit dataframes with specific column names.

Depending on the environmental information required different dates can be 
included, but submitted dataframes must have at least a column titled "Year".
Other date columns can include "Month" and "Day". "Year" and "Day" must be 
numeric, but "Month" can be numeric or character (e.g: 1, "Jan", "1").

tpi: Must contain columns titled: "Tm" and "Binomial". "Tm" contains a 
temperature value, "Binomial" contains species binomial. "TMax" and "TMin" can
also be used as column names containing maximum and minimum temperature values.

IF no month is submitted then the function defualts to using yearly average 
upper and lower thermal tolerance limits for TPI calculation. "Month" can be
specified in the as numeric (1-12) or character ("Jan"-"Dec").

api: Must contain columns titled: "Ar" and "Binomial". "Ar" contains an Aridity
Index value, "Binomial" contains species binomial. "AMax" and "AMin" can
also be used as column names containing maximum and minimum Aridity Index 
values.

IF no month is submitted then the function defaults to using yearly average 
upper and lower Aridity tolerance limits for API calculation. "Month" can be
specified in the as numeric (1-12) or character ("Jan"-"Dec").

#Common Issues:
- misspelled binomials: must be capitalized first letter on genus name followed
by a space then all lowercase species name

- Incorrectly entered data: Aridity Index Values must be positive, and negative
numbers for Ar, AMin, AMax in api function will throw and error. Lat and Lon 
must be in decimal degrees.
